---
title: "02_clean.qmd"
format: html
editor: visual
---

# Clean Data

## Clean HC patient data from sheet1 - dat_load1

```{r}

# Load data 
dat_load1 <- read_tsv(
  file = "~/group_18_project/data/01_dat_load1.tsv",
  show_col_types = FALSE)

# Remove rows that do not correspond to patients and last column called "Notes" 
dat_load1 <- dat_load1 |> 
  slice(1:80) |> select(1:28)
```

### Give shorter names to the variables renaming them

```{r}

# Rename variables 
change_variable_names <- c(
  date_of_birth = "dob",
  grading_system = "ctcae_4_0_hematuria_grade",
  comorbidities = "comorbidities_dm_stroke_pvd_cirrhosis_drug_abuse_alcoholism_tobacco_hx_mi_last_6mo_chf_copd_esrd",
  cancer_type = "type_bladder_prostate_pathologic_clinial_staging_of_cancer",
  radiation_type = "type_of_radation",
  radiation_dosage = "dose_of_radation_gy",
  start_end_radiation = "start_date_of_radiation_end_date_of_radiation",
  number_of_admissions = "number_of_admissions_including_er_visits_requiring_urology_consultation",
  number_of_icu_admits = "number_of_icu_admits_with_hemorrhgic_cystits",
  radiation_completion_to_first_hc = "time_passed_since_completion_of_radiation_treatment_months_and_first_incidence_hc",
  number_of_cystoscopies = "number_of_cystoscopies_with_clot_evacuation_and_or_fulguration_of_bleeders",
  formalin_instillation = "number_of_times_requiring_formalin_instillation",
  alum_instillation = "number_of_times_requiring_alum_instillation",
  silver_nit_instillation = "number_of_times_requiring_silver_nitrate_instillation",
  amicar_instillation = "number_of_times_requiring_amicar_instillation",
  hyperbaric_o2 = "number_times_requiring_hyperbaric_o2_tx_days",
  surgery_for_urinary_diversion = "surgery_for_urinary_diversion_required_if_yes_what_type",
  number_transfusions = "number_of_units_of_p_rb_cs_tranfused",
  hospitalization_complications = "complications_during_hospitalization_uti_pyelonephritis_urosepsis_hydronephrosis_aki_other"
)


dat_load1 <- rename(dat_load1, 
                    any_of(change_variable_names))
```

### Correct mistakes in some observations

```{r}

dat_load1 <- dat_load1 |> 
  mutate(
  
    # Make variable type as Date
    date_of_birth = ymd(date_of_birth),
    
    # ethnicity - make all names lower case, change AA to african american
    ethnicity = ifelse(ethnicity == "AA", "african_american", tolower(ethnicity)),
    
    # hospital_site - remove spaces and separate different hospitals by comas, correct names 
    hospital_site = str_replace_all(hospital_site, " ", ""),
    hospital_site = str_replace_all(hospital_site, "/", ","),
    hospital_site = str_replace(hospital_site, "Harperr", "Harper"),
    hospital_site = str_replace(hospital_site, "Haper", "Harper"),
   
    # radiation_dosage - remove characters "Gy" make the observatons as num
    radiation_dosage =  str_replace_all(radiation_dosage, "Gy", ""),
    radiation_dosage = round(as.numeric(radiation_dosage),2),
    
    # comorbidities add "," to allow separation by ",", str_replace did not work
    comorbidities = ifelse(row_number() == 2, "HTN, A. fib, DM, CHF, tobacco", comorbidities),

    # los_for_each_admission_days replace spaces with comas 
    los_for_each_admission_days = str_replace_all(los_for_each_admission_days, "\\s+", " "),
    los_for_each_admission_days = str_replace_all(los_for_each_admission_days, ", ", " "),
     los_for_each_admission_days = str_replace_all(los_for_each_admission_days, " ", ","),
        los_for_each_admission_days = ifelse(row_number() == 20, "1", los_for_each_admission_days),
    los_for_each_admission_days = ifelse(row_number() == 77, "3,1,4,5,1", los_for_each_admission_days),

    # hospitalization_complications add coma to replace 'which' and 'and' 
    hospitalization_complications = str_replace_all(hospitalization_complications, "and", ","),
    hospitalization_complications = str_replace_all(hospitalization_complications, "which", ","),
    
    # surgery_for_urinary_diversion values to yes or no only
    surgery_for_urinary_diversion = ifelse(tolower(surgery_for_urinary_diversion) == "no", "no", "yes")
    )
    
```

### Clean cancer_type making new columns cancer_site and cancer_stage

```{r}
dat_load1 <- dat_load1 |>
  mutate(
    cancer_site = case_when(
      grepl("prostate", tolower(cancer_type)) & grepl("bladder", tolower(cancer_type)) ~       "prostate,bladder",
      grepl("prostate", tolower(cancer_type)) ~ "prostate",
      grepl("bladder", tolower(cancer_type)) ~ "bladder",
      TRUE ~ NA_character_),
    cancer_stage = str_extract(cancer_type, "(?i)T\\d++[a-z]|Gleason \\d+")) |>
  select(-cancer_type)
```

### Changing names in radiation_type so they fit with the ones in dat_load2 and dat_load3

```{r}

dat_load1 <- dat_load1 |>
  mutate(
    radiation_type = str_replace(radiation_type,"Adjuvant EBRT s/p prostatectomy for a pT4N0" , "EBRT,adjuvant"),
    radiation_type = str_replace(radiation_type,"Adjuvant EBRT s/p prostatectomy with positive LNs" , "EBR,adjuvant"),
    radiation_type = str_replace(radiation_type,"EBRT then seeds" , "EBRT"),
    radiation_type = str_replace(radiation_type,"EBRT, seeds" , "EBRT"),
    radiation_type = str_replace(radiation_type,"EBRT, seeds, intravescial chemo" , "EBRT"),
    radiation_type = str_replace(radiation_type,"EBRT, TURP" , "EBRT"),
    radiation_type = str_replace(radiation_type,"EBRT,brachytherapy,TURBT" , "EBRT,brachytherapy"),
    radiation_type = str_replace(radiation_type,"EBRT,seed implant, lupron" , "EBRT"),
    radiation_type = str_replace(radiation_type,"IMRT boost" , "IMRT"),
    radiation_type = str_replace(radiation_type,"IMRT/tomo" , "IMRT"),
    radiation_type = str_replace(radiation_type,"Mixed neutron and Photon IMRT" , "mixed_neutron_photon_IMRT"),
    radiation_type = str_replace(radiation_type,"mixed neutron/photon EBRT" , "mixed_neutron_photon_EBRT"),
    radiation_type = str_replace(radiation_type,"Neutron Irradiation and Intensity-Modulated Photon Therapy (IMRT)" , "IMRT"),
    radiation_type = str_replace(radiation_type,"Neutron/photon IMRT" , "mixed_neutron_photon_IMRT"),
    radiation_type = str_replace(radiation_type,"photon EBRT" , "mixed_neutron_photon_EBRT"),
    radiation_type = str_replace(radiation_type,"photon/neutron mixed EBRT" , "mixed_neutron_photon_EBRT"),
    radiation_type = str_replace(radiation_type,"radical prostactomey, Adjuvant radiation" , "adjuvant"),
    radiation_type = str_replace(radiation_type,"Salvage EBRT" , "EBRT,salvage"),
    radiation_type = str_replace(radiation_type,tolower("Neutron and mixed_neutron_photon_EBRT"),"mixed_neutron_photon_EBRT"),
    # The previous method str_replace did not work with these observations
    radiation_type = ifelse(row_number() == 19, "IMRT", radiation_type),
    radiation_type = ifelse(row_number() == 22, "mixed_neutron_photon_IMRT", radiation_type),
    radiation_type = ifelse(row_number() == 39, "EBRT,adjuvant", radiation_type),
    radiation_type = ifelse(row_number() == 43, "EBRT,salvage", radiation_type),
    radiation_type = ifelse(row_number() == 57, "IMRT", radiation_type),
    radiation_type = ifelse(row_number() == 59, "mixed_neutron_photon_EBRT", radiation_type),
    radiation_type = ifelse(row_number() == 65, "EBRT", radiation_type),
    radiation_type = ifelse(row_number() == 68, "mixed_neutron_photon_EBRT", radiation_type),
    radiation_type = ifelse(row_number() == 72, "adjuvant", radiation_type),
    radiation_type = ifelse(row_number() == 79, "mixed_neutron_photon_EBRT,adjuvant", radiation_type),
    radiation_type = tolower(radiation_type)
    )
```

### Remove start_end_radiation and dates_of_admissions (too messy)

```{r}
dat_load1 <- dat_load1 |>
  select(-start_end_radiation, -dates_of_admissions)
```

### Move columns

```{r}

dat_load1 <- dat_load1 |>
  relocate(c(cancer_site,cancer_stage), .before=grading_system)
```

### Save the cleaned data as a TSV file in data folder

```{r}
# Save clean data from sheet 1
write.table(x = dat_load1,
            file = "~/group_18_project/data/02_dat_clean1.tsv", 
            sep = "\t", 
            quote = TRUE,
            row.names = FALSE)
```

## Clean patient_data for sheets 2 and 3 - dat_load2 and dat_load3

```{r}

# load data 
dat_load2 <- read_tsv(
  file = "~/group_18_project/data/01_dat_load2.tsv",
  show_col_types = FALSE)

dat_load3 <- read_tsv(
  file = "~/group_18_project/data/01_dat_load3.tsv",
  show_col_types = FALSE)
```

```{r}
# Remove column from the variables and columns that don't exist in patient_info2
dat_load2 <- dat_load2 |> 
  select(-rads_for_prostate_primary, 
         -rads_bladder_primary, 
         -developed_bladder_first,
         -developed_prostate_first)

# Change variable in patient_info1 blader_ca to blader as in patient_info2
dat_load2 <- dat_load2 |>
  rename(bladder = "bladder_ca")
 
# Merge patient_info1 and patient_info2
combined_dat_load2_load3 <- bind_rows(dat_load2, dat_load3)

# Make rename hc as hc_incidence and remove no_hc. If obs=1 patient had hc if obs=0 patient didn't have hc 
combined_dat_load2_load3 <- combined_dat_load2_load3 |> 
  rename("hc_incidence" = "hc") |>
  select(-"no_hc")

# Rename the identifier fin to hospital_id 
combined_dat_load2_load3 <- combined_dat_load2_load3|>
  rename(hospital_id = "fin") 

# Rename 
combined_dat_load2_load3 <- combined_dat_load2_load3|>
  rename(prostate_cancer = "prostate", bladder_cancer = "bladder") 

# Combine types of treatment into one column called treatment
combined_dat_load2_load3 <- combined_dat_load2_load3 |>
  pivot_longer(cols = ebrt:brachytherapy,
              names_to = "treatment", 
              values_to = "values",
              values_drop_na = TRUE) |>
  select(-values)

# Replace NA and Xs for 0 and 1s but yes
combined_dat_load2_load3 <- combined_dat_load2_load3 |> 
   mutate_all(~ ifelse(is.na(.) | tolower(.) == "na", "no", 
                       ifelse(tolower(.) == "x", "yes", .))) 

```

### Save the cleaned data as a TSV file in data folder

```{r}

# Save combined clean data from sheet 2 and sheet 3
write.table(x = combined_dat_load2_load3,
            file = "~/group_18_project/data/02_dat_clean2.tsv", 
            sep = "\t", 
            quote = TRUE,
            row.names = FALSE)
```
