---
title: "03_describe.qmd"
format: html
editor: visual
---

# Describe

Add a short intro to the dataset from article (what is dat_clean 1 and dat_clean 2 about) - hc patients/all patients in the study

## Load dat_clean from data folder

```{r}
dat_clean1 <- read_tsv(
  file = "~/group_18_project/data/02_dat_clean1.tsv",
  show_col_types = FALSE)

dat_clean2 <- read_tsv(
  file = "~/group_18_project/data/02_dat_clean2.tsv",
  show_col_types = FALSE)
```

## Inspect the dat_clean1

```{r}

dat_clean1 |> head(10)
```

Describe the data from dat_clean1 here (total number of patients in dataset and the more important variables we want to study)

```{r}
dat_clean1 |>
  mutate(hematuria_grade = factor(hematuria_grade)) |>
  table1(x = formula(~ age_at_radiation + ethnicity + radiation_type | hematuria_grade),
         data = _)
```

## Inspect the dat_clean2

```{r}

dat_clean2 |> head(10)
```

Describe the data from dat_clean2 here (total number of patients in dataset and the more important variables we want to study)

```{r}
dat_clean2 |>
  table1(x = formula(~ treatment + bladder_cancer | prostate_cancer),
         data = _)
```

```{}
```

```{}
```
