---
title: "04_augment.qmd"
format: html
editor: visual
---

# Augment

## Load .tsv files from data folder

```{r}
dat_clean1 <- read_tsv(
  file = "~/group_18_project/data/02_dat_clean1.tsv",
  show_col_types = FALSE)

# Only if we augment dat_clean2
#dat_clean2 <- read_tsv(
  #file = "~/group_18_project/data/02_dat_clean2.tsv",
  #show_col_types = FALSE)
```

### Create new variables and remove variables that won't be used further

#### Data Wrangling I: BMI

Compute the Body Mass Index (BMI) using the Height and Weight columns. The BMI will be added in a new column named: 'BMI'.

```{r}
dat_clean1 <- dat_clean1 |>     
  mutate(height_m = height_cm/100,     
         BMI = weight_kg/(height_m*2)) |>          
  relocate(BMI, .after = age_at_radiation) 
```

#### Data Wrangling II: num_comorbidities

Counting the number of comorbidities presented bye each patient. The number is added in a new column named: 'num_comorbidities'.

If comorbidities appears like 'NA' a 0 is assigned to 'num_comorbidities'.

```{r}
dat_clean1 <- dat_clean1 |>     
  mutate(num_comorbidities = ifelse(is.na(comorbidities), 0,                      
                                    str_count(comorbidities, pattern = ",") + 1)) |>
  relocate(num_comorbidities, .after = comorbidities) 
```

#### Data Wrangling III: num_hosp_complications

Compute the sum fo hospitalization complications and add is as a new columns called: 'num_hosp_complications'

```{r}
dat_clean1 <- dat_clean1 |>   
  mutate(num_hosp_complications = ifelse(is.na(hospitalization_complications), 0,     
                                         str_count(hospitalization_complications, pattern = ",") + 1)) |>        
  relocate(num_hosp_complications, .after = hospitalization_complications) 
```

#### Data Wrangling IV: num_days_admited

Sum the lenght of stay per admission (los_for_each_admission) to a total of days admitted and add to a new column called: 'num_days_admitted'

```{r}
# Function to sum values in a string  
# Function to sum values in a string
sum_values <- function(string) {
  if (string != "") {
    values <- as.numeric(unlist(strsplit(string, ",")))
    return(sum(values))
  } else {
    return(0)
  }
}

dat_clean1 <- dat_clean1 |>       
  mutate(los_for_each_admission = ifelse(is.na(los_for_each_admission), 0,        
                                         los_for_each_admission),     
         num_days_admitted = sapply(los_for_each_admission, sum_values)) |>     
  relocate(num_days_admitted, .after = los_for_each_admission) 
```

#### Data Wrangling V: cancer_type

Clean cancer_type making new columns cancer_type and cancer_stage

```{r}

dat_clean1 <- dat_clean1 |>
  mutate(
    cancer_type = case_when(
      grepl("prostate", tolower(clinical_cancer)) & grepl("bladder", tolower(clinical_cancer)) ~ "both",
      grepl("prostate", tolower(clinical_cancer)) ~ "prostate",
      grepl("bladder", tolower(clinical_cancer)) ~ "bladder",
      TRUE ~ NA_character_),
    cancer_stage = str_extract(clinical_cancer, "(?i)T\\d++[a-z]|Gleason \\d+")) |>
  select(-clinical_cancer)
  
dat_clean1 <- dat_clean1 |>  
  relocate(c(cancer_type,cancer_stage), .before=hematuria_grade)
```

#### Remove data variables that will not be used further and correct variable placing

```{r}
dat_clean1 <- dat_clean1 |>   
  select(
    -height_m,      
    -height_cm,    
    -weight_kg,     
    -comorbidities,     
    -hospitalization_complications,    
    -los_for_each_admission)  

dat_clean1 <- dat_clean1 |> relocate(radiation_completion_to_first_hc, .after = radiation_duration) 
```

### Save the augmented data as 03_dat_aug.tsv in data folder

```{r}

write.table(x = dat_clean1,
            file = "~/group_18_project/data/03_dat_aug1.tsv", 
            sep = "\t", 
            quote = TRUE,
            row.names = FALSE)

# Only if we add something to dat_clean2
#write.table(x = dat_clean2,
            #file = "~/group_18_project/data/03_dat_aug2.tsv", 
            #sep = "\t", 
            #quote = TRUE,
            #row.names = FALSE)
```
