---
title: "05_analysis_1.qmd"
format: html
editor: visual
---

```{r warning=FALSE}
clean_data2 <- read_tsv(
  file = "~/group_18_project/data/02_dat_clean2.tsv",
  show_col_types = FALSE)
```

Analysis 1

```{r}


plot1 <- clean_data2 |> 
  count(treatment, hc_incidence) |>
  group_by(treatment) |>
  mutate(freq = n / sum(n)) |>
  ggplot(aes(x = treatment, 
           y = freq,
           fill = hc_incidence)) +  
  geom_col(position = position_dodge(
    preserve = "single"),
    colour = "black",
    alpha = 0.5) +
  geom_hline(yintercept = 0) +
  labs(x = "Treatment Received",    
       y = "Proportion",              
       fill = "Radiation-Induced Hemorrhagic Cystitis"     
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 45,
                                   hjust = 1,
                                   vjust = 1.1))

plot1
```

Linear Regression:

```{r}
# Read data from the TSV file
data <- read.delim('~/group_18_project/data/02_dat_clean1.tsv')

#Create the plot
plot_lm <- clean_data1 |> 
  na.omit(clean_data1) |> 
  ggplot(aes(x=radiation_dosage, y=BMI)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  theme_minimal(base_size = 16) +
  labs(x = "Radiation Dose",    
       y = "BMI",              
       title = "Linear Regression between BMI and radiation dose"     
  ) +
  theme(plot.title = element_text(hjust = 0.5, size = 16))  # Adjust title position and size

# Perform linear regression
model <- lm(radiation_dosage ~ BMI, data = clean_data1)

# Print regression summary
summary(model)

#Print plot
print(plot_lm)

```

<<<<<<< HEAD
Calculate percentage

```{r}
# Assuming 'data' is your data frame with columns 'radiation_type' and 'radiation_dosage'
# Define dose categories based on your research or expert consultation
low_dose_threshold <- 50
high_dose_threshold <- 95

# Categorize doses using mutate
clean_data1 <- clean_data1 %>%
  mutate(dose_category = cut(radiation_dosage, breaks = c(-Inf, low_dose_threshold, high_dose_threshold, Inf),
                             labels = c('Low Dose', 'Medium Dose', 'High Dose')))

# Calculate percentages for each type of radiation using summarise
result <- clean_data1 %>%
  group_by(radiation_type, dose_category) %>%
  summarise(Percentage = n() / nrow(clean_data1) * 100) %>%
  ungroup()  # Ungroup the data

# Create a grouped bar plot
ggplot(result, aes(x = radiation_type, y = Percentage, fill = dose_category)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  labs(title = "Percentage of Radiation Dosage in Each Dose Category by Radiation Type",
       x = "Radiation Type",
       y = "Percentage") +
  scale_fill_manual(values = c('Low Dose' = 'lightblue', 'Medium Dose' = 'lightgreen', 'High Dose' = 'lightcoral', 'NAs'='darkgrey')) +
  theme_classic(base_size = 8) + 
  theme(axis.text.x = element_text(angle = 65, hjust = 1, vjust = 1))

=======
  
Plot for comparison of the average time passed since the completion of radiation treatment and the first incident of HC

```{r warning=FALSE}
clean_data1 <- read_tsv(
  file = "~/group_18_project/data/03_dat_aug.tsv",
  show_col_types = FALSE)
```

```{r}
plot2_data <- clean_data1 |>
  select(new_radiation_type, radiation_completion_to_first_hc) |>
  mutate(TimePassed = as.numeric(radiation_completion_to_first_hc)) |>
  filter(!is.na(TimePassed)) |> 
  filter(!is.na(new_radiation_type)) 

result <- plot2_data |>
  group_by(new_radiation_type) |>
  summarize(
    AvgTimePassed = mean(TimePassed),
    MedianTimePassed = median(TimePassed),
    TotalObservations = n())

plot2 <- 
  ggplot(result, aes(x = new_radiation_type, y = AvgTimePassed, fill = new_radiation_type)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(title = "Comparison of Average Time Since Radiation and Time to HC",
       x = "Type of Radiation",
       y = "Average Time Since Radiation (months)") +
  theme_minimal()

print(plot2)
```

```{r}
# Assuming you have a data frame named clean_data1 with columns "radiation_type" and "radiation_dosage"

# Calculate the percentage of patients who received low, medium, and high doses for each radiation type
percentage_df <- clean_data1 |> 
  group_by(radiation_type)  |> 
  summarize(
    Low = sum(radiation_dosage < 50) / n() * 100,
    Medium = sum(radiation_dosage >= 50 & radiation_dosage < 70) / n() * 100,
    High = sum(radiation_dosage >= 70) / n() * 100
  )

# Print the resulting data frame
print(percentage_df)
>>>>>>> b104fcd48b6cac26af67ca5f50224c5755376559

```
